### 2010-08-02 ms/main.tex
shock tube works, leap frog integrator, summation and continuity density implemented

This work aimes at the numerical simluation of high enthalpy flow in
porosities. The subject is of interest, as CITATION
NEEDED (Prof Hornung) has discovered porous surfaces to inhibit/block a
certain form of instabilities in hypersonic flows over a body/cone. A numerical
simulation of this configuration is desired to furnish data for a detailed
analysis of the acting phenomena. The simulation method of choice for the first approach
to this configuration is called Smoothed Particle Hydrodynamics (SPH). Although
there has been no known application of SPH to this precise/concrete case yet/before,
SPH is adopted, as it promises a crucial advantage in the way boundary
conditions are implemented. And for the treatment of porosities,
boundaries/walls represent a non negligible part of the problem.

SPH designates a meshfree method for numerical simulation of particle
ensembles (respectively matter that can be modelled as an ensemble of
particles as it is the case for fluids).
Thereby,the term particle ensembles not only refers
to fluids (as the name of the method would suggest) but also to solids. In
fact nowadays SPH is employed to simulate problems in different fields,
such as body elasticity and fracture,liquid flows and gas
dynamics~\cite{Monaghan2005}.
The initial developement of the method dates back to 1977, where Gingold and
Monaghan tempted to compute an astrophysical gas dynamics problem with a new, more
efficient method~\cite{Gingold1977}. Independently and simultaneously, work in
this field was also carried out by Lucy who came up with the same
idea~\cite{Lucy1977}.
ADVANTAGES???DISADVANTAGES???

@@

The detailed derivation of those equations will not be shown here, as the
interest  only lies in describing the principle ideas behind the SPH
method. To resume, this are the following ones: A (field) function can be
described as an integral interpolation over the function itself and a kernel
function (which has to be chosen in an appropriate way). A further
approximation consists in transforming the integral over the domain into a
summation over the discrete elements constituing this domain (the particles). This allows to express both
the field function and its spacial derivatives dependent exclusively on discrete values
of the fieldfunction itself (and of the kernel function, which is known as well as
its derivatvies). By finally choosing an appropriate kernel function with
compact support, the summation over all particles in the domain can be reduced
to a summation over those particles within the support domain of the particle
in question.

As already mentionned above, the SPH formulation of the equations of fluid
dynamics is obtained by applying the above principles and conducting different
transformations to the original (exact) equations. Depending on the
transformations used, the final SPH equations can assume a different form. In
the following, a common form for the change rates of density, velocity and
internal energy based on the EULER EQUATIONS (i.e. without viscosity) is
listed\cite{Monaghan2005,Liu2003} (perhaps also list NSG-form, and
explain why Euler is of interest as well (for 1DSPH code, perhaps later real
code including viscosity???)

@@

Another point that has to be taken into account for SPH simulations, especially
such including shock phenomena, is arificial viscosity\cite{Monaghan2005}. It
prevents particles from penetrating one into an other in zones with high
compression rates and also generally stabilizes a numerical
allgorithm.
Artificial viscosity for numerical simulation of shock phenomena
in general was first introduced by \cite{vonNeumann1950}. They suggested to
introduce an artificial dissipation to give the shock a thickness which
allowed it to be calculated numerically (in the cited example with finite difference
scheme) without introduction of pre-/after shock boundary conditions.    ...(see article)
The first use of artificial viscosity specifically for the mesh free SPH goes
back to Lucy\cite{Lucy1977}. He suggested an artificial bulk viscosity to
prevent a slow buildup of acoustic energy by integration errors in zones with
low density respectively low particle presence. 
Finally, the form of artificial viscosity most commomly used in SPH
literature\cite{Liu2003} is the one proposed by Monaghan and
Gingold\cite{Monaghan1983}. They stated that the use of one of the artificial
viscosities cited above (Neumann Richtmyer (adapted to SPH) or bulk
viscosity) on shock problems would lead to either excessive particle
oscillation or excessive smearing of the shock front. The expression for this
artificial viscosity $\Pi_{ab}$, which is added to the pressure term of
equation \ref{eq:VCR_Euler}, is:

@@

$v_{ab}$ stands for $v_a-v_b$, $r_{ab}$ for $r_a-r_b$, $c_{ab}$ is the
averaged sound speed of the two particles in question $1/2(c_a+c_b)$. The same
applies for $\rho_{ab}(=1/2(\rho_a+\rho_b))$ and $h_{ab}=(1/2(h_a+h_b))$. IT
HAS TO BE MENTIONED SOMEWHERE (but probably not right here) THAT: the use of
the averaged smoothing length $h_{ab}$ is a way to symmetrize particle
interactions when h is spatially variable. In this case a particle $a$ could be
located in the support domain of a particle $b$ but not vice versa. This would
lead to an assymetric mutual contribution of forces. To avoid this, using the
averaged smoothing length between two particles is an appropriate remedy\cite{Liu2003}.
Including artificial viscosity, the SPH form of the Euler momentum equation becomes\cite{Monaghan2005}

@@

\section{Implementation of an SPH code}
This section presents some particular features of an SPH code. Emphasis is
placed on these features which are finally implemented in one of the
two SPH codes designed and modified during this work.
Points of interest are the implementation of different boundary conditions in
SPH, a closer description of the kernel function, as well as different methods/allgorithms
to perform the search for the interacting particles. Furthermore different
equations of state are introduced, depending on the nature of the flow problem
to be considered(compressible/incompressible). Finally some words on numerical
integration schemes that have proven to be reliable for SPH applications.

\subsection{boundary condition implementation}
\label{sec:boundaryConditionImplementation}
For particles at the edge of the calculation domain, the summation interpolation does not
give exact results, as the integral (or better the sum) is truncated due to
the simple fact that outside the domain there are no
particles.\cite{Liu2003}(find better citation...) This efect is called the
problem of article deficiencythe particle deficiency
problem.\cite{Liu2003}. In some cases there is no need to consider this
problem, as the areas close to the edges are not of interest and due to
limited simulation time there is no time for the error to propagate into the
actual zone of interest. The 1D shock tube simulation described further below
(PERHAPS LINK TO SECTION)is an example for such a case.  
But for other applications the edges play an important role or the domain has
to be rendered periodocally. So there is a need to define specific boundary
conditions. This can be accomplished by a special boundary treatment, which
often includes so called virtual particles or ghost
particles\cite{Liu2003}. The first ones to come up with this idea were
\cite{LIBERSKY1993}. Virtual particles are used to model solid walls by
placing them inside the solid body, mirrored at the surface. For each real
particle whose support domain interfres with the wall, one such particle is
built. Virtual particles have the same properties as their real counterpart
except for the velocity. Depending on the type of boundary condition they are
supposed to represent, different velocities are assigned to them.\cite{Hu2006}

@@

Some properties of Kernel functions (smoothing functions) and notions related
to them have allready been mentionned in the section
\ref{sec:BasicsSPH}. In principal every function can act as Kernel function
when it respects a set of criteria complsed by \cite{Liu2003}.
The most important ones are:

@@

In the work of \cite{Fulk1996} a mesure of merit for Kernels is developed and
they came to the result, that in 1D no kernel performed significantly better
than the cubic spline.
This is why for the 1D SPH code described in section (WHATEVER) the cubic
spline kernel is employed.

\subsection{neighboring particle search methods}
\label{sec:NNPS}
To calculate the change rates in \ref{eq:DCR_Euler}-\ref{eq:ECR_Euler} a
summation over all paricles has to be performed. However, as we defined kernel
functions with a compact support, only particles within a certain distance from
the particle in question (within its support domain) contribute to the
interpolation of its values, and thus interact. Finding the interacting
particles is one of the key features in an SPH implementation, as an effective
interaction search method (often also called nearest neighboring particle (NNP)
search method in SPH literature) can considerably speed up/accelerate
computation time.

@@

\begin{figure}[h]
  \centering
      \includegraphics[width=1.0\textwidth]{Graphics/general_structure_SPH}
  \caption{basic structure of an SPH code}
  \label{fig:BasicSphCode}
\end{figure}

>>>>

This work aims at the numerical simulation of high enthalpy flow in
porosities. The subject is of interest, as CITATION
NEEDED (Prof Hornung) has discovered porous surfaces to inhibit/block a
certain form of instabilities in hypersonic flows over a body/cone. A numerical
simulation of this configuration is desired to furnish data for a detailed
analysis of the acting phenomena. The simulation method of choice for the first approach
to this configuration is called Smoothed Particle Hydrodynamics (SPH). Although
there has been no known application of SPH to this precise/concrete case yet/before,
SPH is adopted, as it promises a crucial advantage in the way boundary
conditions are implemented. And for the treatment of porosities,
boundaries/walls represent a non negligible part of the problem.

SPH designates a meshfree method for numerical simulation of particle
ensembles (respectively matter that can be modelled as an ensemble of
particles as it is the case for fluids).
Thereby,the term particle ensembles not only refers
to fluids (as the name of the method would suggest) but also to solids. In
fact nowadays SPH is employed to simulate problems in different fields,
such as body elasticity and fracture,liquid flows and gas
dynamics~\cite{Monaghan2005}.
The initial development of the method dates back to 1977, where Gingold and
Monaghan tempted to compute an astrophysical gas dynamics problem with a new, more
efficient method~\cite{Gingold1977}. Independently and simultaneously, work in
this field was also carried out by Lucy who came up with the same
idea~\cite{Lucy1977}.
ADVANTAGES???DISADVANTAGES???

@@

The detailed derivation of those equations will not be shown here, as the
interest  only lies in describing the principle ideas behind the SPH
method. To resume, this are the following ones: A (field) function can be
described as an integral interpolation over the function itself and a kernel
function (which has to be chosen in an appropriate way). A further
approximation consists in transforming the integral over the domain into a
summation over the discrete elements constituting this domain (the particles). This allows to express both
the field function and its spacial derivatives dependent exclusively on discrete values
of the fieldfunction itself (and of the kernel function, which is known as well as
its derivatives). By finally choosing an appropriate kernel function with
compact support, the summation over all particles in the domain can be reduced
to a summation over those particles within the support domain of the particle
in question.

As already mentioned above, the SPH formulation of the equations of fluid
dynamics is obtained by applying the above principles and conducting different
transformations to the original (exact) equations. Depending on the
transformations used, the final SPH equations can assume a different form. In
the following, a common form for the change rates of density, velocity and
internal energy based on the EULER EQUATIONS (i.e. without viscosity) is
listed\cite{Monaghan2005,Liu2003} (perhaps also list NSG-form, and
explain why Euler is of interest as well (for 1DSPH code, perhaps later real
code including viscosity???)

@@

Another point that has to be taken into account for SPH simulations, especially
such including shock phenomena, is artificial viscosity\cite{Monaghan2005}. It
prevents particles from penetrating one into an other in zones with high
compression rates and also generally stabilizes a numerical
algorithm.
Artificial viscosity for numerical simulation of shock phenomena
in general was first introduced by \cite{vonNeumann1950}. They suggested to
introduce an artificial dissipation to give the shock a thickness which
allowed it to be calculated numerically (in the cited example with finite difference
scheme) without introduction of pre-/after shock boundary conditions.    ...(see article)
The first use of artificial viscosity specifically for the mesh free SPH goes
back to Lucy\cite{Lucy1977}. He suggested an artificial bulk viscosity to
prevent a slow buildup of acoustic energy by integration errors in zones with
low density respectively low particle presence. 
Finally, the form of artificial viscosity most commonly used in SPH
literature\cite{Liu2003} is the one proposed by Monaghan and
Gingold\cite{Monaghan1983}. They stated that the use of one of the artificial
viscosities cited above (Neumann Richtmyer (adapted to SPH) or bulk
viscosity) on shock problems would lead to either excessive particle
oscillation or excessive smearing of the shock front. The expression for this
artificial viscosity $\Pi_{ab}$, which is added to the pressure term of
equation \ref{eq:VCR_Euler}, is:

@@

$v_{ab}$ stands for $v_a-v_b$, $r_{ab}$ for $r_a-r_b$, $c_{ab}$ is the
averaged sound speed of the two particles in question $1/2(c_a+c_b)$. The same
applies for $\rho_{ab}(=1/2(\rho_a+\rho_b))$ and $h_{ab}=(1/2(h_a+h_b))$. IT
HAS TO BE MENTIONED SOMEWHERE (but probably not right here) THAT: the use of
the averaged smoothing length $h_{ab}$ is a way to symmetrize particle
interactions when h is spatially variable. In this case a particle $a$ could be
located in the support domain of a particle $b$ but not vice versa. This would
lead to an asymmetric mutual contribution of forces. To avoid this, using the
averaged smoothing length between two particles is an appropriate remedy\cite{Liu2003}.
Including artificial viscosity, the SPH form of the Euler momentum equation becomes\cite{Monaghan2005}

@@

\section{Implementation of an SPH code}
This section presents some particular features of an SPH code. Emphasis is
placed on these features which are finally implemented in one of the
two SPH codes designed and modified during this work.
Points of interest are the implementation of different boundary conditions in
SPH, a closer description of the kernel function, as well as different methods/algorithms
to perform the search for the interacting particles. Furthermore different
equations of state are introduced, depending on the nature of the flow problem
to be considered(compressible/incompressible). Finally some words on numerical
integration schemes that have proven to be reliable for SPH applications.

\subsection{boundary condition implementation}
\label{sec:boundaryConditionImplementation}
For particles at the edge of the calculation domain, the summation interpolation does not
give exact results, as the integral (or better the sum) is truncated due to
the simple fact that outside the domain there are no
particles.\cite{Liu2003}(find better citation...) This effect is called the
problem of article deficiencythe particle deficiency
problem.\cite{Liu2003}. In some cases there is no need to consider this
problem, as the areas close to the edges are not of interest and due to
limited simulation time there is no time for the error to propagate into the
actual zone of interest. The 1D shock tube simulation described further below
(PERHAPS LINK TO SECTION)is an example for such a case.  
But for other applications the edges play an important role or the domain has
to be rendered periodically. So there is a need to define specific boundary
conditions. This can be accomplished by a special boundary treatment, which
often includes so called virtual particles or ghost
particles\cite{Liu2003}. The first ones to come up with this idea were
\cite{LIBERSKY1993}. Virtual particles are used to model solid walls by
placing them inside the solid body, mirrored at the surface. For each real
particle whose support domain interfres with the wall, one such particle is
built. Virtual particles have the same properties as their real counterpart
except for the velocity. Depending on the type of boundary condition they are
supposed to represent, different velocities are assigned to them.\cite{Hu2006}

@@

Some properties of Kernel functions (smoothing functions) and notions related
to them have already been mentioned in the section
\ref{sec:BasicsSPH}. In principal every function can act as Kernel function
when it respects a set of criteria composed by \cite{Liu2003}.
The most important ones are:

@@

In the work of \cite{Fulk1996} a measure of merit for Kernels is developed and
they came to the result, that in 1D no kernel performed significantly better
than the cubic spline.
This is why for the 1D SPH code described in section (WHATEVER) the cubic
spline kernel is employed.

\subsection{neighboring particle search methods}
\label{sec:NNPS}
To calculate the change rates in \ref{eq:DCR_Euler}-\ref{eq:ECR_Euler} a
summation over all particles has to be performed. However, as we defined kernel
functions with a compact support, only particles within a certain distance from
the particle in question (within its support domain) contribute to the
interpolation of its values, and thus interact. Finding the interacting
particles is one of the key features in an SPH implementation, as an effective
interaction search method (often also called nearest neighboring particle (NNP)
search method in SPH literature) can considerably speed up/accelerate
computation time.

@@

\begin{figure}[h]
  \centering
     % \includegraphics[width=1.0\textwidth]{Graphics/general_structure_SPH}
  \caption{basic structure of an SPH code}
  \label{fig:BasicSphCode}
\end{figure}
